Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-11-28 inst/manuscript/manuscript.tex
more updates

The novel Laplace approach implemented in \pkg{spmodel} formally
maximizes a hierarchical GLM likelihood
\citep{lee1996hierarchical, wood2017generalized}, making
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. Next, we describe a brief overview
of the approach and how it can be used for several primary data analysis
tasks \citep{tredennick2021practical} like model comparison, parameter
estimation, inference, model diagnostics, and prediction.

@@

The fitted values (\code{.fitted}) can be returned on either the link
(\(\hat{\mathbf{w}}\)) or response (\(f^{-1}(\hat{\mathbf{w}})\)) scale
and the residuals (\code{.resid}) can be deviance, Pearson, or response
residuals. The default fitted values are on the link scale and the
default residuals are deviance residuals. Also returned by
\code{augment()} are the leverage (\code{.hat}), Cook's distance
(\code{.cooksd}), and standardized residuals (\code{.std.resid})
described in Section\(~\)\ref{sec:diagnostics}. A benefit of using
\code{augment()} when \code{data} is an \code{sf} object is that the
output is also an \code{sf} object, which makes it straightforward to
create spatial diagnostic plots (Figure\(~\)\ref{fig-sp-diagnostic}).
Standard \proglang{R} helpers (e.g., \code{fitted()},
\code{residuals()}) are also available to extract model diagnostics from
the model object.

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-2.png}
\caption{Moose presence model diagnostics, including leverage values (left) and standardized residuals (right).}
\label{fig-sp-diagnostic}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
\caption{Moose presence probability fitted values and predictions. Fitted values are represented by circles and predictions by triangles.}
\label{fig-moose-fit}
\end{figure}

Predictions on the response scale are visualized alongside the fitted
values (\(f^{-1}(\hat{\mathbf{w}})\)) in Figure\(~\)\ref{fig-moose-fit}.
Prediction intervals for the probability of moose presence (on the link
scale) are returned by supplying \code{interval}:

@@

Throughout the remainder of this section, we briefly highlight some
additional \pkg{spmodel} capabilities for SPGLMs. In
Section\(~\)\ref{sec-moose-count}, we fit Poisson and negative binomial
models with and without geometric anisotropy for the point-referenced
moose count data. In Section\(~\)\ref{sec-lake}, we fit a Gamma model to
the point-referenced lake conductivity data, showing how to fit a model
with a partition factor, perform a spatial analysis of variance (ANOVA),
and estimate contrasts for models with interactions. In
Section\(~\)\ref{sec-seal}, we fit a binomial model to the areal harbor
seal trend data with a nonspatial random effect. Finally in
Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
data, which can be treated as point-referenced or areal, and use maximum
likelihood to compare two models with different explanatory variables.
Table\(~\)\ref{tab:additional} outlines, for each application, the
section number, data set, family (i.e., response distribution), geometry
type (point-referenced or areal), and additional \pkg{spmodel} features
highlighted.

@@

The spatial covariance is strongest in a northwest-southeast direction
and weakest in the northeast-southwest direction
(Figure\(~\)\ref{fig-tropy}), which is intuitive given the similar
patterns in moose counts from Figure\(~\)\ref{fig-moose-data-count}.

\begin{figure}[h]
\centering
\includegraphics[width = 1\linewidth]{figures/figure-7.png}
\caption{Level curves of equal spatial covariance for the negative binomial moose count models. The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of spatial covariance.  In the isotropic level curve (left), spatial covariance decays equally in all directions. In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
\label{fig-tropy}
\end{figure}

@@

The \code{seal} data in \pkg{spmodel} contains harbor seal abundance
trends for two different harbor seal stocks (genetically distinct
populations). While the \code{moose} and \code{lake} data were
point-referenced, the \code{seal} data are areal. Each polygon in the
\code{seal} data represents a distinct harbor seal haulout region
(Figure\(~\)\ref{fig-seal}). A haulout region is an area of coastal
rocks that harbor seals go to rest, molt, and give birth.

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-8.png}
\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed log seal trends (right).}
\label{fig-seal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
\caption{Proportion of voter turnout in Texas for the 1980 presidential election. Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
\label{fig-texas}
\end{figure}

The \code{texas} data in \pkg{spmodel} contains voter turnout data for
Texas counties in the 1980 United States Presidential Election
\citep{bivand2024spdata}. The data are point-referenced, with polygon
centroids representing the spatial location of each county
(Figure\(~\)\ref{fig-texas}). Beta regression is a GLM used to model
rate and proportion data in the (0, 1) interval
\citep{ferrari2004beta, cribari2010beta}. We model voter turnout rates
as a function of mean log income of county residents using an SPGLM
assuming a beta distributed response variable:

@@

Alternatively, we could use an autoregressive model to fit the model,
constructing a neighborhood matrix by assuming centroids within
\code{cutoff} of one another are neighbors:

@@

According to AIC, the SPGLM for point-referenced data is preferred:

@@

The default estimation method in \pkg{spmodel} for SPGLMs is restricted
maximum likelihood (REML), while maximum likelihood (ML) can also be
used. A benefit of benefit of REML is that it can yield unbiased
estimates of covariance parameters \citep{cressie1993asymptotic}, but a
drawback is that likelihood-based statistics are only valid for model
comparison when the models have the same explanatory variable and fixed
effect structure (because the error contrasts used to construct the REML
likelihood change based on \(\mathbf{X}\) and \(\boldsymbol{\beta}\)).
In contrast, ML estimators are generally biased for covariance
parameters, though in practice this bias tends to be small. Moreover,
when using ML, likelihood-based comparisons are valid for models having
different explanatory variable and fixed effect structures. Using ML, we
can evaluate the significance of log income on voter turnout using a
likelihood ratio test:

\begin{CodeChunk}
\begin{CodeInput}
R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
R> spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
R> anova(spbeta_full_ml, spbeta_red_ml)
\end{CodeInput}
\begin{CodeOutput}
Likelihood Ratio Test

Response: turnout
                                Df   Chi2 Pr(>Chi2)    
spbeta_red_ml vs spbeta_full_ml  1 23.155 1.494e-06 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> AIC(spbeta_full_ml, spbeta_red_ml)
\end{CodeInput}
\begin{CodeOutput}
               df       AIC
spbeta_full_ml  7 -31.25900
spbeta_red_ml   6 -10.10354
\end{CodeOutput}
\end{CodeChunk}

@@

SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
approximation that simultaneously marginalizes over the latent (i.e.,
unobserved) random effects and the fixed effects. \pkg{spmodel}'s
\code{spglm()} (for point-referenced data) and \code{spgautor()} (for
areal data) fit SPGLMs that are similar in structure and syntax as base
\proglang{R}'s \code{glm()} function, easing the transition from GLMs to
SPGLMs for practitioners. The \code{spglm()} and \code{spgautor()}
functions support six response distributions for binary, count, and
skewed data and 20 spatial covariance functions. \pkg{spmodel} has a
suite of tools for data visualization, inference, model diagnostics, and
prediction, providing a framework that can be used for all stages of a
data analysis. There are many additional \pkg{spmodel} features that are
not covered here, including fitting multiple models simultaneously,
fixing spatial covariance and dispersion parameters at known values,
fitting models to large non-Gaussian data having thousands of
observations via spatial indexing \citep{ver2023indexing}, incorporating
spatial dependence in machine learning (e.g., random forests;
\citet{breiman2001random}), simulating spatially dependent data (e.g.,
\code{spbinom()}, \code{sprpois()}, etc.), and more. Further details are
provided by \url{https://CRAN.R-project.org/package=spmodel} and links
therein.

@@

All writing and code associated with this manuscript is available for
viewing and download on GitHub at
\url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
part of the \pkg{spmodel} {R} package available for download from CRAN
at \url{https://CRAN.R-project.org/package=spmodel}. Results were
obtained using {R} 4.4.0 with the \pkg{spmodel} 0.11.1 package. Figures
were created using the {ggplot2} 3.5.1 package
\citep{wickham2016ggplot2} and base {R}.

>>>>

The novel Laplace approach implemented in \pkg{spmodel} formally
maximizes a hierarchical GLM likelihood
\citep{lee1996hierarchical, wood2017generalized}, making
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not directly available for
quasi-likelihood \citep{wedderburn1974quasi, breslow1993approximate} or
pseudo-likelihood approaches \citep{wolfinger1993generalized}, which
only specify the first two moments of a distribution. Next, we describe
a brief overview of the approach and how it can be used for several
primary data analysis tasks \citep{tredennick2021practical} like model
comparison, parameter estimation, inference, model diagnostics, and
prediction.

@@

The fitted values (\code{.fitted}) can be returned on either the link
(\(\hat{\mathbf{w}}\)) or response (\(f^{-1}(\hat{\mathbf{w}})\)) scale
and the residuals (\code{.resid}) can be deviance, Pearson, or response
residuals. The default fitted values are on the link scale and the
default residuals are deviance residuals. Also returned by
\code{augment()} are the leverage (\code{.hat}), Cook's distance
(\code{.cooksd}), and standardized residuals (\code{.std.resid}). A
benefit of using \code{augment()} when \code{data} is an \code{sf}
object is that the output is also an \code{sf} object, which makes it
straightforward to create spatial diagnostic plots
(Figure\(~\)\ref{fig-sp-diagnostic}). Standard \proglang{R} helpers
(e.g., \code{fitted()}, \code{residuals()}) are also available to
extract model diagnostics from the model object.

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-4.png}
\caption{Moose presence model diagnostics, including leverage values (left) and standardized residuals (right).}
\label{fig-sp-diagnostic}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
\caption{Moose presence probability fitted values ($f^{-1}(\hat{\mathbf{w}})$) and predictions. Fitted values are represented by circles and predictions by triangles.}
\label{fig-moose-fit}
\end{figure}

Moose are far more common in the eastern and southwestern portion of the
domain (Figure\(~\)\ref{fig-moose-fit}). Prediction intervals for the
probability of moose presence (on the link scale) are returned by
supplying \code{interval}:

@@

Throughout the remainder of this section, we briefly highlight some
additional \pkg{spmodel} capabilities for SPGLMs. In
Section\(~\)\ref{sec-moose-count}, we fit Poisson and negative binomial
models with and without geometric anisotropy for the point-referenced
moose count data. In Section\(~\)\ref{sec-lake}, we fit a Gamma model to
the point-referenced lake conductivity data, showing how to fit a model
with a partition factor, perform a spatial analysis of variance (ANOVA),
and estimate contrasts for models with interactions. In
Section\(~\)\ref{sec-seal}, we fit a binomial model to the areal harbor
seal trend data with a nonspatial random effect. Finally in
Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
data, which can be modeled using point-referenced or areal support, and
use maximum likelihood to compare two models with different explanatory
variables. Table\(~\)\ref{tab:additional} outlines, for each
application, the section number, data set, family (i.e., response
distribution), geometry type (point-referenced or areal support), and
additional \pkg{spmodel} features highlighted.

@@

From Figure\(~\)\ref{fig-tropy}, the spatial covariance is strongest in
a northwest-southeast direction (2.729 radians) and weakest in the
northeast-southwest direction (0.413 radians), which is intuitive given
the similar patterns in moose counts from
Figure\(~\)\ref{fig-moose-data-count}.

\begin{figure}[h]
\centering
\includegraphics[width = 1\linewidth]{figures/figure-9.png}
\caption{Level curves of equal spatial covariance for the negative binomial moose count models. The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of spatial covariance.  In the isotropic level curve (left), spatial covariance decays equally in all directions. In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction (0.413 radians) and slowest in the northwest-southeast direction (2.729 radians).}
\label{fig-tropy}
\end{figure}

@@

The \code{seal} data in \pkg{spmodel} contains harbor seal abundance
trends for two different harbor seal stocks (genetically distinct
populations). While the \code{moose} and \code{lake} data have
point-referenced support, the \code{seal} data have areal support. Each
polygon in the \code{seal} data represents a distinct harbor seal
haulout region (Figure\(~\)\ref{fig-seal}). A haulout region is an area
of coastal rocks that harbor seals go to rest, molt, and give birth.

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-10.png}
\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed log seal trends (right).}
\label{fig-seal}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-11.png}
\caption{Proportion of voter turnout in Texas for the 1980 presidential election. Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
\label{fig-texas}
\end{figure}

The \code{texas} data in \pkg{spmodel} contains voter turnout data for
Texas counties in the 1980 United States Presidential Election
\citep{bivand2024spdata}. The data have point-referenced support, with
polygon centroids representing the spatial location of each county
(Figure\(~\)\ref{fig-texas}). Beta regression is a GLM used to model
rate and proportion data in the (0, 1) interval
\citep{ferrari2004beta, cribari2010beta}. We model voter turnout rates
as a function of mean log income of county residents using a
geostatistical SPGLM assuming a beta distributed response variable:

@@

Alternatively, we could fit an autoregressive SPGLM by assuming the data
have areal support and constructing a neighborhood matrix that treats
counties as neighbors if the distance between their centroids is less
than \code{cutoff}:

@@

According to AIC, the geostatistical SPGLM:

@@

The default estimation method in \pkg{spmodel} for SPGLMs is restricted
maximum likelihood (REML), while maximum likelihood (ML) can also be
used. A benefit of benefit of REML is that it can yield unbiased
estimates of covariance parameters \citep{cressie1993asymptotic}. A
drawback of REML is that likelihood-based statistics are often
considered invalid when the models have different explanatory variable
or fixed effect structures \citep{wolfinger1993covariance}, though
\citet{gurka2006selecting} provides some evidence to the contrary. In
contrast to REML estimators, ML estimators are generally biased for
covariance parameters, though in practice this bias tends to be small.
Moreover, when using ML, likelihood-based comparisons are valid for
models with different explanatory variable or fixed effect structures.
Using ML, we can evaluate the significance of log income on voter
turnout using a likelihood ratio test (for nested models):

\begin{CodeChunk}
\begin{CodeInput}
R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
R> spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
R> anova(spbeta_full_ml, spbeta_reduced_ml)
\end{CodeInput}
\begin{CodeOutput}
Likelihood Ratio Test

Response: turnout
                                    Df   Chi2 Pr(>Chi2)    
spbeta_reduced_ml vs spbeta_full_ml  1 23.155 1.494e-06 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> AIC(spbeta_full_ml, spbeta_reduced_ml)
\end{CodeInput}
\begin{CodeOutput}
                  df       AIC
spbeta_full_ml     7 -31.25900
spbeta_reduced_ml  6 -10.10354
\end{CodeOutput}
\end{CodeChunk}

@@

SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
approximation that simultaneously marginalizes over the latent (i.e.,
unobserved) random effects and the fixed effects. \pkg{spmodel}'s
\code{spglm()} (for point-referenced support) and \code{spgautor()} (for
areal support) fit SPGLMs that are similar in structure and syntax as
base \proglang{R}'s \code{glm()} function, easing the transition for
practitioners from GLMs to SPGLMs. The \code{spglm()} and
\code{spgautor()} functions support six response distributions for
binary, count, and skewed data and 20 spatial covariance functions.
\pkg{spmodel} has a suite of tools for data visualization, inference,
model diagnostics, and prediction, providing a framework that can be
used for all stages of a data analysis. There are many additional
\pkg{spmodel} features that are not covered here, including fitting
multiple models simultaneously, fixing spatial covariance and dispersion
parameters at known values, fitting models to large non-Gaussian data
having thousands of observations via spatial indexing
\citep{ver2023indexing}, incorporating spatial dependence in machine
learning (e.g., random forests; \citet{breiman2001random}), simulating
spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()},
etc.), and more. Further details are provided by
\url{https://CRAN.R-project.org/package=spmodel} and links therein.

@@

All writing and code associated with this manuscript is available for
viewing and download on GitHub at
\url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
part of the \pkg{spmodel} {R} package available for download from CRAN
at \url{https://CRAN.R-project.org/package=spmodel}.
